---
title: "Statistical Research Skills -- Assignment 2"
author: "Lorenzo Mella (UUN: s1566023)"
output: pdf_document
---


# Description of the problem

Kernel density estimation (KDE) models the distribution under a criterion of smoothness of the density. Under the heuristic principle that datapoints in a sample generated form a distribution with density $f(x)$ tautologically accumulate with high probability in high-density regions, we associate to each datapoint $x_i$ a scaled kernel $K_h(x-x_i)$. The average of these functions hopefully provides a good approximation to the original density.

The choice of kernel is dictated by numerical requirements: if the kernel has unbounded support, for instance, every summand must be computed in $x$, to evaluate the density $\hat{f}(x)$.

It is also dictated by smoothness considerations: if we expect the data to be generate from a smooth density, it would be more sensible to use graded kernels such as a Gaussian or the Epanechnikov kernel. A rectangular kernel would be more appropriate for piecewise constant densities.


## Kernel density estimator

Problematic choices:

- bandwidth
- kernel shape

```{r}

```

## Histogram estimator

Problematic choices:

- number of bins

## Mixture model estimator

Problematic parameter:

- conditional distributions
- number of conditional distributions


# Experimental setting

## Densities that represent a wide testing range of possibilities

```{r}
# Define densities and RNGs for the distributions to be tested

# Normal (unimodal base case)
ddensity_1 = function(x) dnorm(x, 0, 3)
rdensity_1 = function(n) rnorm(n, 0, 3)
# Student t (unimodal with heavier tails)
ddensity_2 = function(x) dt(x / 3, df = 1)
rdensity_2 = function(n) 3 * rt(n, df = 1)
# Uniform (discontinuous density with a dense modes)
ddensity_3 = function(x) dunif(x, 0, 3)
rdensity_3 = function(n) runif(n, 0, 3)
# Gamma (unimodal, positively skewed)
ddensity_4 = function(x) dgamma(x, 3, 1)
rdensity_4 = function(n) rgamma(n, 3, 1)
# Mixture of distributions
# (trimodal, with many challenging features)
ddensity_5 = function(x) {
  0.5 * dnorm(x, 0, 1) + 0.3 * dgamma(8 - x, 2, 1.5) + 0.2 * dunif(x, 3, 4.5)
}
rdensity_5 = function(n) {
  num_samples = rmultinom(1, n, c(0.5, 0.3, 0.2))
  sample(c(rnorm(num_samples[1], 0, 1),
           8 - rgamma(num_samples[2], 2, 1.5),
           runif(num_samples[3], 3, 4.5)))
}
```

## One shot experiments

- Graphs
- Discussion

```{r}
# A density function obtained from histogram data returned by hist

hist_density = function(hist_data, x) {
  # Find the bin where x lies: its index is the maximum number
  # of breaks less or equal to x. Return the histogram normalised
  # frequency of that bin
  sapply(x, FUN = function(y) {
    if (y < hist_data$breaks[1] |
          y >= hist_data$breaks[length(hist_data$breaks)]) {
      return(0)
    } else {
      return(hist_data$density[sum(hist_data$breaks - y <= 0)])
    }
  })
}
O```


```{r}
max_samples = 250
# Generate data from the first density
xx = rdensity_5(max_samples)

plot(density(xx))

normalmix_density = function(x, mix_output) {
  sapply(x, function(y) sum(mix_output$lambda * dnorm(y, mix_output$mu, mix_output$sigma)))
}

mixres = normalmixEM(xx, k = 2)

plot(seq(0, 10, by = 0.1), normalmix_density(seq(0, 10, by = 0.1), mix_output = mixres), type = "l")
```


## Monte Carlo simulation

- ISEs for sample sizes 250, 500, 1000

This essentially means that 

- Discussion
